``Interestingness'' in Mathematical Theory Exploration

### 2019-06-14 PhDThesis/bucketing.tex
Bucketing changes in thesis

Figure~\ref{fig:survival} shows a Kaplan-Meier survival plot of \quickspec{}
running times, when given inputs containing different numbers of definitions. Many runs
finish quickly, with the remainder occupying a ``long tail'' which we cut off
after 300 seconds\footnote{Chosen based on preliminary experiments, which showed
  little difference in survival between 300 seconds and 1 hour}.

@@

Figure~\ref{fig:tailsize} shows that the number of definitions in the input is
linearly correlated with the proportion of timeouts, except for the very
smallest inputs (which are more constrained by the sampling procedure). One
hypothesis to explain this is the existence of ``toxic'' definitions, whose
presence in the input always leads to the exploration failing (with larger
inputs being more likely to have sampled a toxic definition).

@@

This does appear to be the case, as Figure~\ref{fig:proportions} shows that five
definitions appeared \emph{only} in failing inputs; these are named
\texttt{mul3}, \texttt{mul3acc}, \texttt{mult2}, \texttt{op} and \texttt{qexp};
their definitions appear in Figure~\ref{fig:faildefs}. All of these are
functions of Peano-encoded natural numbers (\texttt{Nat}), and they cause
exploration to time out by either exhausting the RAM with too many expressions
to explore, % TODO{2019-04-27} CAN QSPEC2 AVOID THIS?
or by generating such deeply-nested outputs that comparing them takes too long.
% TODO{2019-04-27} CAN WE PUT TIME AND MEMORY LIMITS ON EACH ATTEMPTED EVALUATION?

@@

\begin{itemize}
\item Avoid ``toxic'' definitions: those producing exponentially large outputs,
  and/or taking many arguments (leading to a large number of combinations). In
  particular, exposing the accumulator argument of tail-recursive functions can
  make them harder to explore.
\item Smaller sets are faster to explore than larger sets, but the
\end{itemize}

It also seems prudent for theory exploration tools to limit time and memory
usage at the level of individual evaluations: these can abort evaluation of
toxic definitions, whilst still being high enough to allow the majority of
non-toxic definitions to be unhindered.

@@

Our final criterion concerns the ``quality'' of the selected subsets. This
depends crucially on how we define whether a property is ``interesting''. We can
use the same ground truth definition as our Theory Exploration Benchmark: a
property is ``interesting'' if it appears in a particular corpus (in our case
TIP~\cite{claessen2015tip}), and ``uninteresting'' if not. Whilst na\"ive, this
approach allows the criterion to be quantified: the quality of the selected
subsets is proportion of the whole signature's interesting properties that also
apply to the subsets.

Since signature selection is largely independent of which tool we choose to
explore the resulting subsets, we do not require that such exploration actually
\emph{finds} all of the interesting properties; only that we have not
\emph{prevented} them from being found. For example, we may be given a signature
containing addition and multiplication functions and may consider commutativity
and distributivity properties to be ``interesting''. Selecting a subset
containing both would be high quality, since all of these properties are
\emph{available} to find, even if our chosen tool doesn't spot them. In
contrast, splitting addition and multiplication into \emph{separate} subsets is
lower quality, since this prevents \emph{any} tool finding the distributivity
property (since it involves both definitions).

To understand the quality impact of limiting ourselves to such subsets, we
applied four signature selection methods to signatures sampled from the Theory
Exploration Benchmark. Signatures varied in size from a single definition up to
100 definitions and 100 signatures of each size were sampled. Each method was
run 19 times on each signature: splitting into between 1 and 20 subsets.

All four methods place each definition in exactly one subset, to focus on the
effect of splitting apart definitions. The tested methods are:

@@

The optimal and pessimal methods are unrealistic, since they require access to
the ground truth (which only exists for benchmark problems), but set upper and
lower bounds on the impact of separating a signature's definitions. These are
implemented via constraint solving, whose exponential running time became
infeasible for signatures containing more than 10 definitions. To avoid extreme
solutions, such as all-but-one subsets being singletons, subset cardinality
could only differ by at most one.

The pseudorandom and clustering implementations are more realistic, since they
do not use any information from the ground truth. Cardinality limits were also
not imposed for these methods.

% TODO: Insert graph of hash vs MLSpec on 1-100

% TODO: Insert graph of 1-10 with all

>>>>

Figure~\ref{fig:survival} shows a Kaplan-Meier survival plot of \quickspec{}
running times, when given inputs containing different numbers of definitions.
Many runs finish quickly, with the remainder occupying a ``long tail'' which
didn't finish within our 300 second timeout\footnote{Chosen based on preliminary
  experiments, which showed little difference in survival between 300 seconds
  and 1 hour}.

@@

The height of these ``tails'' is linearly correlated with the number of
definitions in the signature, as shown in Figure~\ref{fig:tailsize}. One
hypothesis to explain this is the existence of ``toxic'' definitions, whose
presence in the input always leads to the exploration failing (with larger
inputs being more likely to have sampled a toxic definition).

@@

To investigate the plausibility of such definitions, the ratio of successful and
timed-out runs is plotted for each name in Figure~\ref{fig:proportions}; showing
that five definitions appeared \emph{only} in failing inputs, and seem likely to
be ``toxic''. These definitions are named \texttt{mul3}, \texttt{mul3acc},
\texttt{mult2}, \texttt{op} and \texttt{qexp}; their definitions appear in
Figure~\ref{fig:faildefs}. All of these are functions of Peano-encoded natural
numbers (\texttt{Nat}), and they cause exploration to time out by either
exhausting the RAM with too many expressions to explore, % TODO{2019-04-27} CAN QSPEC2 AVOID THIS?
or by generating such deeply-nested outputs that comparing them takes too long.
% TODO{2019-04-27} CAN WE PUT TIME AND MEMORY LIMITS ON EACH ATTEMPTED EVALUATION?

@@

\begin{itemize}
\item Avoid ``toxic'' definitions: those producing exponentially large outputs,
  and/or taking many arguments (leading to a large number of combinations). In
  particular, tail-recursive functions which expose their accumulator argument
  are harder to explore.
\item Smaller signatures are faster to explore, so should be preferred if
  possible.
\end{itemize}

It also seems prudent for theory exploration tools to limit time and memory
usage, both at the global level (to abort if it becomes clear we are in the
``long tail'') and at the level of individual evaluations. In particular such
fine-grained limits can abort evaluation of toxic definitions, whilst still
being lenient enough to allow the majority of non-toxic definitions to be
explored unhindered.

@@

Our final criterion concerns the ``quality'' of the selected subsets. This
depends crucially on how we define whether a property is ``interesting''. We can
use the same ground truth definition as our Theory Exploration Benchmark: a
property is ``interesting'' if it appears in a particular corpus (in our case
TIP~\cite{claessen2015tip}), and ``uninteresting'' if not. Whilst na\"ive, this
approach allows the criterion to be quantified: the quality of the selected
subsets is the proportion of the whole signature's interesting properties that
also apply to the subsets taken individually.

Since signature selection is largely independent of which tool we choose to
explore the resulting subsets, we do not require that such exploration actually
\emph{finds} all of the interesting properties; only that we have not
\emph{prevented} them from being found. For example, we may be given a signature
containing addition and multiplication functions and may consider commutativity
and distributivity properties to be ``interesting''. A subset containing both
would be a high quality selection, since all of these properties are
\emph{available} to find, even if our chosen tool doesn't spot them. In
contrast, splitting addition and multiplication into \emph{separate} subsets is
lower quality, since this stops \emph{any} tool from finding the distributivity
property (since it involves both definitions).

To understand the quality impact of limiting ourselves to such subsets, we
applied four signature selection methods to signatures sampled from the Theory
Exploration Benchmark. Signatures varied in size from a single definition up to
100; for each size, we sampled 100 signatures.

All four methods were forced to place each definition in exactly one subset, to
focus on the effect of splitting apart definitions rather than e.g. duplicating
definitions across several subsets or dropping some entirely. The tested methods
are:

@@

The optimal and pessimal methods are unrealistic, since they require access to
the ground truth (which only exists for benchmark problems), but set upper and
lower bounds on the impact of separating a signature's definitions. These are
implemented via constraint solving, whose exponential running time became
infeasible for signatures containing more than 10 definitions. To avoid extreme
solutions, such as placing all definitions in one subset and leaving the others
empty, we required that all of the selected subsets are non-empty and their
cardinality differs by at most one. As a consequence, these methods cannot
divide signatures of size $N$ into more than $N$ subsets (since there are not
enough definitions to populate every subset without duplicating).

The pseudorandom and clustering implementations are more realistic, since they
do not use any information from the ground truth. These methods may also leave
some of their subsets empty (since, unlike the optimal and pessimal methods,
they are less prone to extremes).

\begin{figure}
  % \input{images/bounds.pgf}
  \includegraphics[width=\textwidth]{images/bounds.png}
  \caption{The average quality (proportion of ground-truth properties available)
    when splitting signatures sampled using the Theory Exploration Benchmark
    into subsets. Each plot shows division into a different number of subsets.
    The shaded area is bounded by the optimal and pessimal methods, which are
    can only produce $N$ subsets for signatures of size $\geq N$. Grey lines
    show pseudorandom selection, for comparison; these extend into smaller
    signature sizes since empty subsets are permitted.}
  \label{fig:bounds}
\end{figure}

Figure~\ref{fig:bounds} shows the impact of selection (specifically, dividing
definitions between subsets, with no skipping or duplication) on the proportion
of interesting properties that remain available for subsequent theory
exploration tools to discover. The shaded region shows the difference between
(average) quality, with optimal and pessimal selections starting out equal (when
dividing signatures of $N$ definitions into $N$ subsets, where there is no
choice other than putting each definition on its own) but quickly diverging as
choices become available. Note that the optimal and pessimal methods are
constrained to avoid empty or unequally-distributed sets, so they cannot divide
up a signature of size $N$ into more than $N$ subsets. Pseudorandom selection is
shown as a grey line, and although the cardinality of its subsets is not
constrained, it nevertheless appears consistently between the bounds.

Two trends are visible in these charts: firstly the bounded region trends
upwards as the signature size increases, which is promising for the application
of signature selection to break apart large signatures. Secondly the upper bound
reduces as the number of subsets increases: this is to be expected as more
definitions must be separated to produce the extra sets, but it demonstrates the
tradeoff between quality and subset size. The distance between the pseudorandom
and optimal results shows the potential gains to be made by smarter signature
selection methods.

\begin{figure}
  % \input{images/bucketing}
  \includegraphics[width=\textwidth]{images/bucketing.png}
  \caption{Comparison of subset quality when selected by recurrent clustering
    (solid lines) versus pseudorandom selection (dashed lines). Darker lines
    show division into a greater number of subsets (between 1 and 20). The solid
    and dashed lines follow each other closely, indicating that clustering
    performs no better than random at signature selection.}
  \label{fig:bucketing}
\end{figure}

The signature selection problem is superficially similar to the
\emph{clustering} problem in machine learning: grouping data points based on
their similarity across multiple dimensions. To investigate this potential
relationship we created a signature selection method based on a \emph{recurrent
  clustering} algorithm for Haskell code: this flattens syntax trees into
vectors, substitutes fixed numbers for keywords and recurses to determine
suitable numbers for referenced expressions.

The results of this method are shown in Figure~\ref{fig:bucketing}, along with
the pseudorandom method as a control. The results of clustering turn out to be
no better than random; in fact the curves show surprisingly similar patterns.
